Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2015-07-24 MSH.tex
Second draft

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle is probably the source for some kimberlitic eruptions, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. The ringwoodite stability field requires that the kimberlite must have been sourced from at least ca. 500 km depth. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition. Composition is a key variable in determining melt density, viscosity and solid-solid-melt dihedral angles, which together with the temperature and equivalent properties of the solid control the time- and lengthscales on which melt can separate from the melt. They also govern the extent to which this melt can become channelised, reducing interaction with its surroundings and increasing ascent rates. Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their water concentrations must be estimated indirectly. Compositions of partial hydrous melts have been estimated using a mixture of mass balance and defocused-beam electron probe measurements, which to provide Mg:Si ratios and a rough estimate of water content from deficits in totals. Such measurements are conducted on melt pockets \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011} or melts isolated within diamond traps \citep{MSUP2007}. Unfortunately, these techniques are accompanied by significant uncertainties, especially in terms of water content. An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt \emph{ex-situ} (when much of the water has been lost). The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content. For example, in the MgO-SiO$_2$ system, maximum water contents in a solid starting material are provided by a mixture of brucite and quartz. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.

In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

@@

Multianvil experiments were performed in the 5000 tonne press at the Bayerisches Geoinstitut (BGI). Cr-doped MgO octahedral multianvil assemblies with 18 mm edge length were used (Figure \ref{fig:assembly}). Two capsules were loaded into each assembly, with the open ends of the chambers facing each other and separated by six 0.05 mm thick Pt$_{90}$Rh$_{10}$ or Au foils. The assemblies were compressed to 13 GPa over four hours between eight tungsten carbide cubes with trucations of edge-length 11 mm. Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}. The assemblies were then resistively heated via the stepped LaCrO$_3$ heater. The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly. To avoid water loss, higher temperature runs were heated for a shorter duration. The experiments were quenched by cutting power to the heater. The assemblies were then decompressed to room pressure over 1000 minutes.

@@

\section{Modelling water solubility in melts}

Oxygen in MgO-SiO$_2$-H$_2$O melts can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group. The general equation describing the reaction between melt species is
\begin{equation}
0.5 \textrm{H}_2\textrm{O} + 0.5 \textrm{O}_{br}^{2-} \rightleftharpoons \textrm{OH}_{tm}^-
\label{eqn:speciation}
\end{equation}

where the subscripts br and tm respectively refer to bridging and terminal oxygens. The equilibrium between these species can be described with an equilibrium constant $K$ \cite[e.g.][]{Stolper1982}
\begin{equation}
K = \frac{X_{\textrm{OH}_{tm}^-}}{\sqrt{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} } }
\label{eqn:equilibrium_constant}
\end{equation}

The equilibrium constant is related to the energy $\Delta G^{\circ}$ required to protonate a bridging oxygen
\begin{equation}
K = \exp\left(\frac{-\Delta G^{\circ}}{RT}\right)
\end{equation}

@@

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers. No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions. For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$. In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G^{\circ} = a + b\,T
\end{equation}

\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K=\infty$, $\Delta G^{\circ}=-\infty$, $r$ maximised). Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.

To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the aforementioned models by allowing non-ideality between the two endmembers. The excess term $\Delta G^{\circ}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.

In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
\begin{equation}
RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
\end{equation}
where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.

In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$. For MgO melt, only the reference Helmholtz free energy needed to be adjusted. We note that the melting point of MgO at high pressure is rather uncertain. In the present model \citep{DKS2013}, MgO melts at $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. Experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. We note that if the present model is incorrect, it is probably the calculated entropy of melting which is at fault. A higher melting temperature would therefore imply a lower entropy of melting, such that the values in $G_{\textit{melt}}$ - $G_{\textit{solid}}$ at low temperatures will have smaller errors, than would otherwise be expected from such large melting point uncertainties. For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa \citep{ZLGHF1993}. The volume at 1 bar is fit to experimental data. The entropy of melting at the coesite-stishovite-melt triple point ($\sim$40 J/K/mol) was kept the same. This value is identical to the value derived from the Clapeyron slope of the stishovite liquidus \cite{SL1995}. We note that this slope is much steeper than that recently proposed by \cite{Millotetal2015}. This difference is not the result of incompatibility between experimental results at different pressures \citep{SL1995, LAM1983, Millotetal2015}, but instead the result of using a 2-parameter Simon's equation to fit the rather abrupt change in dT/dP implied by the experiments at 20--40 GPa.

A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions. The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.

@@

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
Experimental results for the system MgO-H$_2$O are shown in Figure \ref{fig:MH}. The periclase-brucite-water invariant at 13 GPa is well constrained at 62--66 mol\% H$_2$O and $\sim$1200 $^{\circ}$C. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. The melting curve and equilibrium with brucite is modelled with the ideal OH$^-$ mixing model of \cite{SS1985}. $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995}. This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system, as observed in the small dT/dX of the brucite liquidus. This flat slope is in excellent agreement with the suggestion that the second critical endpoint in the MgO-H$_2$O system probably lies at ca. 12--13 GPa \citep{MSUP2007}.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/periclase}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (the discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\clearpage
\subsubsection{Mg$_2$SiO$_4$-H$_2$O}
The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}. The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see Figure \ref{fig:ternary}). Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C. This is in excellent agreement with results on the wadsleyite-water system run at 15 GPa \citep{DDFK2005, LSOK2011}. At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite. The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by MgO-deficit in forsterite at these conditions. This MgO deficit may be the result of hydrogen incorporation by the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}. Chambers with higher water contents have forsterites with similar compositions, but melts which are slightly more SiO$_2$-rich.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt. The preferred fit through the data is given by $G^{\circ} = 75 \cdot (1420-T)$ J/mol.}
  \label{fig:foH}
\end{figure}
\clearpage

>>>>

\section{Introduction}
Hydrous melts have a remarkable influence on Earth's surface and interior. Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes. The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}. This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}. In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage. It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.

Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition. Most importantly, the equilibrium proportion of H$_2$O in melts controls the extent of partial melting for a given bulk H$_2$O content. Together with solid-solid-melt dihedral angles, the extent of partial melting governs the permeability and therefore mobility of melt, and the ability to detect it via body wave seismology and electrical conductivity measurements.  Composition is also a key variable in determining melt density and viscosity. Finally, the change in equilibrium melt compositions with pressure and temperature has a significant influence on melt channelisation, which reduces melt-solid interaction and increases ascent rates.

Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure. Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth. Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O. As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases. As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one. Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.

Constraining melt compositions remains a major challenge in high-pressure experimental petrology. The primary problem is that water-rich melts are unquenchable. During quench, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and sample preparation. The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}. Mass balance alone depends on knowing the volume proportions of solid phases and quench phases, being able to recognise and subtract quench overgrowths where they exist, and knowing the effective ambient-pressure density of the melt. All of these are associated with significant uncertainties. Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit would imply a large water content). However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam. For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).

Bracketing the liquidus requires only that the bulk composition during the experiment is known. This technique therefore seems a good alternative to estimating melt compositions ex-situ. The only problem is that at slab or ambient mantle temperatures, water contents in melts at high pressure are very high. Making up solid starting compositions with more water than a brucite-quartz mix has traditionally been avoided. Existing studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form if the sample chambers are sufficiently large \citep[e.g.][]{HM2012}. Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible. In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

@@

Multianvil experiments were performed in the 5000 tonne press at the Bayerisches Geoinstitut (BGI). Cr-doped MgO octahedral multianvil assemblies with 18 mm edge length were used (Figure \ref{fig:assembly}). Two capsules were loaded into each assembly, with the open ends of the chambers facing each other and separated by six 0.05 mm thick Pt$_{90}$Rh$_{10}$ or Au foils. The assemblies were compressed to 13 GPa over four hours between eight tungsten carbide cubes with trucations of edge-length 11 mm. Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}. The assemblies were then resistively heated via a stepped LaCrO$_3$ heater. The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly. To avoid water loss, higher temperature runs were heated for a shorter duration. The experiments were quenched by cutting power to the heater. The assemblies were then decompressed to room pressure over 1000 minutes.

@@

\section{Modelling water solubility in melts}
\subsection{Background}
Oxygen in neutral MSH melt species can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group. The general equation describing the reaction between melt species is
\begin{equation}
\textrm{H}_2\textrm{O} + \textrm{O}_{br}^{2-} \rightleftharpoons 2 \textrm{OH}_{tm}^-
\label{eqn:speciation}
\end{equation}

where the subscripts br and tm respectively refer to bridging and terminal oxygens. The equilibrium between these species can be described with an equilibrium constant $K$ \cite[e.g.][]{Stolper1982}
\begin{equation}
K_{(1)} = \frac{{X_{\textrm{OH}_{tm}^-}}^2}{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} }
\label{eqn:equilibrium_constant}
\end{equation}

The equilibrium constant is related to the energy $\Delta G_{(1)}$ required to protonate two bridging oxygens with a water molecule
\begin{equation}
K_{(1)} = \exp\left(\frac{-\Delta G_{(1)}}{RT}\right)
\end{equation}

@@

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K_{(1)}=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers. No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K_{(1)}$ in water-rich compositions. For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K_{(1)}<\infty$. In practise, the equilibrium constant $K_{(1)}$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G_{(1)} = a + b\,T
\end{equation}

\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K_{(1)}=\infty$, $\Delta G_{(1)}=-\infty$, $r$ maximised). Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.

To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the aforementioned models by allowing non-ideality between the two endmembers. The excess term $\Delta G_{(1)}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.

These models can be used to calculate melt compositions by equating the gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the component with the composition of that solid in the melt $\mu^s_{melt}$:
\begin{equation}
G^s_{solid} = \mu^s_{melt} = G^s_{melt} - RT \ln a^s_{melt}
\label{eqn:equilibrium}
\end{equation}
In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, stishovite and coesite. A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$. The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.

\subsection{The anhydrous melt model}
Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy. Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour. Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data. First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point. This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations. Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite). It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy. This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}. We notes that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}. Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data. However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs. 27.5 J/K/mol). At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.

The MD-determined entropy of stishovite is significantly lower than the experimentally determined value, so we increased the standard state entropy of the melt by 16.5 J/K/mol. Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa. The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}. Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point. Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower. Therefore, we adjusted the bulk modulus and reference pressure so satisfy this constraint, and the volume at zero pressure. The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \citep{DKS2013}. Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point. The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model. The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point. This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}. We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly. A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.

A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts. The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}. We assumed an ideal entropic contribution to mixing. The pressure dependencies were fit to the melting curves. It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower. However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).

\subsection{The hydrous melt model}

In this study, the anhydrous components in our hydrous melts are normalised to a one-cation basis (i.e. MgO, Mg$_{2/3}$Si$_{1/3}$O$_{4/3}$, Mg$_{1/2}$Si$_{1/2}$O$_{3/2}$, SiO$_2$). We follow \cite{SS1985} in determining the proportions of H$_2$O + O$_{br}$ and OH$_{tm}$  in the melt as a function of the molar proportion of H$_2$O $p_{H_2O}$ as follows:

\begin{eqnarray}
p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} \\
X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
X_{H_2O} = X_O + 2 p' - 1 \\
X_{OH} = 1 - X_O - X_{H_2O}
\end{eqnarray}
\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis). We assume that all bridging oxygens are equally likely to be protonated. The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2). The ideal activities of the anhydrous and water components can then be calculated:
\begin{eqnarray}
a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
\end{eqnarray}

The activity coefficients (i.e. the activity prefactor which provides the non-ideal contribution to mixing; $a = \gamma a_{ideal}$) are equal to one except in the case of the MgO-H$_2$O binary, where a simple Margules model is applied \citep{HM2012}
\begin{equation}
RT \ln \gamma_i = {p_j}^2 \left( 1 - p_j \right) \left( 2 W_{ji} - W_{ij} \right) \\
\end{equation}
\noindent where the subscripts $i$ and $j$ correspond to the components $s$ or H$_2$O. Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs should be divided by 3 and 4 to reduce them to their one-cation equivalents.

@@

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
Experimental results for the system MgO-H$_2$O are shown in Figure \ref{fig:MH}. The periclase-brucite-water invariant at 13 GPa is well constrained at 62--66 mol\% H$_2$O and $\sim$1200 $^{\circ}$C. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. In addition to the thermodynamic models described in the methods section, the activity of H$_2$O at the brucite-periclase-melt triple point is calculated using the thermodynamic dataset of \citep{HP2011}, which includes models for brucite, periclase and water \citep{PS1995}. This provides an additional constraint on the melt model via the equation
\begin{equation}
RT \ln a_{H_2O} =  G_{br} - G_{per} - G_{H_2O}
\end{equation}
The brucite melting curve is obtained by finding the solution to the equation
\begin{equation}
(G_{MgO, melt} + RT \ln a_{MgO}) + (G_{H_2O, HP} + RT \ln a_{H_2O}) = G_{br, HP} - G_{per, HP} + G_{per, SLB}
\end{equation}
\noindent where the right hand side is the chemical potential of Mg(OH)$_2$ in the melt, and the right hand side is the Gibbs free energy of brucite with a correction for the different standard state energies of MgO between the \cite{HP2011} and \cite{SLB2011} datasets. $G_{(1)} = RT \ln K_{(1)} = -75000$ J/mol ($r$=1) provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol). The large positive interaction parameter increases the tendency for melt-fluid immiscibility at the water-rich end of the system. This is reflected in the flat slope of the brucite liquidus, and is in excellent agreement with the suggestion that the second critical endpoint in the MgO-H$_2$O system probably lies at ca. 12--13 GPa \citep{MSUP2007}.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/periclase}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text). Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}. An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
  \label{fig:MH}
\end{figure}

@@

\clearpage
\subsubsection{Mg$_2$SiO$_4$-H$_2$O}
The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}. The model parameters shown are $G_{(1)} = 75 \cdot (1420-T)$ J/mol, $r=4/3$. The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see also Figure \ref{fig:ternary}). Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C. This is in excellent agreement with results on the wadsleyite-water system run at 15 GPa \citep{DDFK2005, LSOK2011}. At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite. The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}. Alternatively it could simply be the result of a small SiO$_2$ deficit in that sample chamber.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/forsterite}
      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text. At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
  \label{fig:foH}
\end{figure}
\clearpage

\subsubsection{MgSiO$_3$-H$_2$O}
The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. Unlike the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}. The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 120 \cdot (1750-T)$ J/mol, $r=3/2$. The disagreement between this model and the experimental observations is small, and could plausibly be due to experimental error. However, given the results for stishovite (see next section), we feel that more complex models are required to explain the depression of melting in the MgSiO$_3$ system.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/enstatite}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
  \label{fig:eoH}
\end{figure}
